The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-10-28 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Finsihed the first draft of the MWA RFI environment article :-)

\section{Introduction}
In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not solve interference from air-based transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}. The consequence of this approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed. It is important to analyze the impact of RFI on a specific instrument to consider the right RFI mitigation approach. Additionally, an increased understanding of the RFI situation will help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

@@

The interferometric arrays LOFAR, PAPER and the Long Wavelength Array (LWA) are all designed to perform 21-cm EoR science, and all observe in approximately the same frequency range as the MWA. Comparing the MWA to these arrays therefore allows the comparison of the observatory site and hardware design with respect to the RFI impact. TODO say something about PAPER and MWA RFI. In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose issues on regular observations even though LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of EoR long-integration observations with LOFAR. They reach near-thermal noise sensitivity, and argue that RFI is not a limitation for further increasing the sensitivity. Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}. Those analyses concluded that with sufficient precautions, such as good receiver design and accurate detection methods, the leaked RFI is weak and averages down similar to Gaussian noise.

TODO paper structure.

\section{Method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast. It also has a library interface, which allows it to be integrated in a pipeline.

@@

\section{Data description}
We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The nights of observations are listed in Table~\ref{tbl:obs-list}.

@@

\section{Detection results}
The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

\section{Comparison with LOFAR}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies. The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while O2012 reports a 3.18\% occupancy for LOFAR.

It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz. As shown in O2012, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions. O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA. In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low declinations. An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations. Because of these differences, it is hard to compare the environments based on detected RFI levels. Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction.

@@

\section{Conclusions \& Discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DVD signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation.

@@

>>>>

\section{Introduction}
In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequence of this approach is that certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed. It is important to analyze the impact of RFI on a specific instrument to consider the right RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

@@

% TODO mention GMRT ? and the Long Wavelength Array (LWA; \citealt{lwa1-ellingson-2013})  ?
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa has been sufficiently clean to integrate sufficiently long to reach a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, which is one magnitude away from expected EoR signals. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose issues on regular observations, even though LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity. Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}. Those analyses conclude that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, the leaked RFI is weak and averages down similar to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

In this paper, we will describe the mitigation efforts implemented for the MWA, and show examples of RFI found and results of the RFI detection. Sect.~\ref{ch:method} describes the approach taken for the MWA, including software, algorithms and computational challenges involved. This strategy is applied to 63~h of MWA data, and these data are described in Sect.~\ref{ch:data-description}. Sect.~\ref{ch:detection-results} presents the RFI that was found in these data, as well as the effectivity of the RFI detection. In Sect.~\ref{ch:lofar-comparison}, these results will be compared to a previously performed LOFAR RFI survey. Finally, conclusions are drawn and discussed in Sect.~\ref{ch:conclusions-and-discussion}.

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast. It also has a library interface, which allows it to be integrated in a pipeline.

@@

\section{Data description} \label{ch:data-description}
We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The nights of observations are listed in Table~\ref{tbl:obs-list}.

@@

\section{Detection results} \label{ch:detection-results}
The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

\section{Comparison with LOFAR} \label{ch:lofar-comparison}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies. The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while O2012 reports a 3.18\% occupancy for LOFAR.

It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz. As shown in O2012, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions. O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA. In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low declinations. An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.

Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI levels. Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction. These values can therefore be interpreted and compared as being the minimum loss of data due to RFI.

@@

\section{Conclusions \& discussion} \label{ch:conclusions-and-discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DVD signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation.

@@

\begin{itemize}
 \item TODO Say something about LEDA (LWA project) and HERA (Parsons' project)
\end{itemize}
